Synthesis of Linear Functional Programs

### 2021-07-16 article/article.tex
WIP: Overview

First, we discuss the use of linear types as specifications, the outline of the
synthesis process, and what that entails for the synthetiser -- ilustrated with
concrete results (Section~\ref{sec:overview}). Then, we present the
algorithm and formal system driving the synthesis
(Section~\ref{sec:formal_system}). Aftwerwards, we describe the architecture 
of our synthetiser named \synname, and examine technical details and challenges of the
implementation (Section~\ref{sec:architecture}). Finally, we evaluate our work
through expressiveness benchmarks (i.e.~can we synth $X$?) and time benchmarks
(i.e.~how fast can we synth a term of type $T$?) (Section~\ref{sec:evaluation}).

% TODO: Why linear types?

%%

%%
%This work explores the synthesis of linear functional
%programs from types based on linear logic propositions (i.e.~linear
%types) by leveraging the Curry-Howard correspondence.

@@

The \synname\ synthetiser combines the features of linear types, recursion,
parametric polymorphism, recursive polymorphic? algebraic data types, and
refinement types.

The \synname\ language is an extension of the linear $\lambda$-calculus.

... in proof search, our programs are correct by construction -- if
we were able to synth it, then it typechecks.

Linear types provide a more certain specification, when compared to simple
traditional types. By requiring some types to be used linearly (consumed) we can
assure, for example, the deconstruction of an ADT. Given the goal $\bang(a \lolli b)
\lolli List\ a \lolli List\ b$, the program $\lambda x. \lambda y. Nil$ wouldn't
be valid. To synthetise a valid program, $List\ a$
would have to be deconstructed, and its constructor arguments used, resulting
in the map function without relying on any additional information:

@@

At the core of the synthesis process, and of the developed synthetiser \synname,
is a \emph{sound} and \emph{complete?} system of focusing for linear logic with
resource management -- a type system consisting of inference rules in sequent
calculus for linear logic connectives; with resource management, and focusing,
as formalized techniques to, respectively, solve resource related
non-determinism, and to eliminate from the proof search don't-care
non-determinism.  Guided solely by this system's formal
representation\ref{sec:formal_system}, we can (and did) create a very robust
synthetiser for a select part of the linear types that's able to synthetise
programs successfully from e.g. the major linear logic equivalences. For
example, given the specification $(A \lolli (B \lolli C)) \lolli (A \tensor B)
\lolli C$, the program $\lambda a .\ \lambda b .$ let $c \tensor d$ = $b$ in $a$
$c$ $d$ (commonly known as \emph{uncurry}) is synthetised.

The \synname\ programming language is naturally born alongside the synthetiser
to create an environment for it. Synth goals are inserted in the program by use
of a \emph{synth} keyword or syntatic \emph{mark}, indicating a program should
be generated for a given, or inferred, type. Synth results are functions in this
language's syntax. Additionally, it opens the possibility of synthetisis with
context, that is, with knowledge of other functions and eventual structures
defined in the same program.

@@

First, after decomposing a recursive ADT, we must consume all constructor
parameters, and since we can't decompose the self type any further because of a
restriction, we should be able to use it from the linear context; even though,
non-restricted types are not to take that course since they weren't unnaturally
restricted, and because it would result in additional non-determinism and
infinite recursion. Second, without any additional rules, an ADT in the linear
context will loop back to the inversion context, and then loop again, going back and
forth between the two contexts for infinity; instead, when focused on an ADT, we should
either instance the goal (provided they're the same type), or switch to
inversion if and only if its decomposition isn't restricted. The three
following rules ensure this:
\begin{mathpar}
    \infer*[right=(adt$\Uparrow$L)]
    {
        (\Rho_C; \Rho_D);\Gamma; \Delta, x{:}T/\Delta'; \Omega \Uparrow\ \vdash M : C
        \and
        T \in \Rho_D
    }
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta'; \Omega, x{:}T \Uparrow\ \vdash M : C}
    \and
    \infer*[right=(adt-init)]
    {  }
    {\Rho; \Gamma; \Delta/\Delta'; x{:}T \Downarrow\ \vdash x : T}
    \and
    \infer*[right=(adt$\Downarrow$L)]
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}T \Uparrow\ \vdash M :
    T \and T \notin \Rho_D}
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}T \Downarrow\ \vdash M : T}
\end{mathpar}
% TODO: Adicionar regra que diz que se for ADT Rec -o A então a construção de
% ADT Rec é logo restrita?? parece que acelera minimamente mas não arranjo um
% exemplo em que seja crucial. pode ser uma coisa que tenha ficado para resolver
% uma coisa antiga que ficou resolvida mais tarde com outra modificação e então
% é agora inutil. Talvez seja demais adicionar à regra

>>>>

First, we discuss the use of linear types as specifications, the outline of the
synthesis process, and what that entails for the synthetiser -- ilustrated with
concrete results (Section~\ref{sec:overview}). Then, we present the
algorithm and formal system driving the synthesis
(Section~\ref{sec:formal_system}). Aftwerwards, we describe the architecture 
of our synthetiser named \synname, and examine technical details and challenges of the
implementation (Section~\ref{sec:architecture}). Finally, we evaluate our work
through expressiveness benchmarks (i.e.~can we synth $X$?) and time benchmarks
(i.e.~how fast can we synth a term of type $T$?) (Section~\ref{sec:evaluation}).
%TODO: And appendice with...

%This work explores the synthesis of linear functional
%programs from types based on linear logic propositions (i.e.~linear
%types) by leveraging the Curry-Howard correspondence.

@@

% TODO: Why linear types?
The \synname\ synthetiser combines the features of linear types with recursion,
parametric polymorphism, recursive polymorphic? algebraic data types, and
refinement types. The synthetiser is built on top of a system of proof-search in
linear logic. Proof-search relates to program synthesis through the Curry-Howard
correspondence~\cite{}, which states that \emph{propositions in a logic} map
directly to \emph{types}, and \emph{proofs} of those propositions to well-typed
\emph{programs} -- finding a proof of a proposition is finding a program with
that type.

The core of the synthesis is a \emph{sound} and \emph{complete} system
consisting of \emph{bottom-up} proof search in linear logic augmented with
resource management~\cite{} and focusing~\cite{} as techniques to, respectively,
solve resource related non-determinism, and eliminate from the proof-search
don't-care non-determinism. A more in depth explanation is in
section~\ref{sec:formal_system}.

With the core in place, we can synthetise programs from specifications that use
canonical linear types ($\lolli, \with, \tensor, 1, \oplus, \bang$), such as
fundamental linear logic equivalences. For example, given the specification $(A
\lolli (B \lolli C)) \lolli (A \tensor B) \lolli C$, the program
\mintinline{haskell}{\a -> \b -> let c*d = b in a c d}
% $\lambda a .\ \lambda b .$ let $c \tensor d$ = $b$ in $a$ $c$ $d$
(commonly known as \emph{uncurry}) is synthetised.

The \synname\ programming language is naturally born alongside the synthetiser
to create an environment for it. Synth goals are inserted in the program by use
of a \emph{synth} keyword or syntatic \emph{mark}, indicating a program should
be generated for a given type. Additionally, it opens the possibility of synthesis with
context, that is, with knowledge of other functions and structures defined in
the same program.

All additions to the synthesis are extensions of the proof-search in linear
logic system -- in the end, the process still consists of constructing a proof.
Because we can only construct valid proofs (a wrong proof isn't a proof), we
know that all produced synthesis results are correct: \textbf{all programs are
correct by construction}, i.e. if we're able to synthetise it, then it typechecks,
and therefore fullfills the specification.

Linear types provide a more certain specification, when compared to simple
traditional types. By requiring some types to be used linearly (consumed) we can
assure, for example, the deconstruction of an ADT. Given the goal $\bang(a \lolli b)
\lolli List\ a \lolli List\ b$, the program \mintinline{haskell}{\x -> \y ->
Nil} wouldn't be valid. To synthetise a valid program,
\mintinline{haskell}{List a}
would have to be deconstructed, and its constructor arguments used, resulting
in the map function without relying on any additional information:

@@

@@

First, after decomposing a recursive ADT, we must consume all constructor
parameters, and since we can't decompose the self type any further because of a
restriction, we should be able to use it from the linear context; even though,
non-restricted types are not to take that course since they weren't unnaturally
restricted, and because it would result in additional non-determinism and
infinite recursion. Second, without any additional rules, an ADT in the linear
context will loop back to the inversion context, and then loop again, going back and
forth between the two contexts for infinity; instead, when focused on an ADT, we should
either instance the goal (provided they're the same type), or switch to
inversion if and only if its decomposition isn't restricted. The three
following rules ensure this:
\begin{mathpar}
    \infer*[right=(adt$\Uparrow$L)]
    {
        (\Rho_C; \Rho_D);\Gamma; \Delta, x{:}T/\Delta'; \Omega \Uparrow\ \vdash M : C
        \and
        T \in \Rho_D
    }
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta'; \Omega, x{:}T \Uparrow\ \vdash M : C}
    \and
    \infer*[right=(adt-init)]
    {  }
    {\Rho; \Gamma; \Delta/\Delta'; x{:}T \Downarrow\ \vdash x : T}
    \and
    \infer*[right=(adt$\Downarrow$L)]
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}T \Uparrow\ \vdash M :
    T \and T \notin \Rho_D}
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}T \Downarrow\ \vdash M : T}
\end{mathpar}
% TODO: Adicionar regra que diz que se for ADT Rec -o A então a construção de
% ADT Rec é logo restrita?? parece que acelera minimamente mas não arranjo um
% exemplo em que seja crucial. pode ser uma coisa que tenha ficado para resolver
% uma coisa antiga que ficou resolvida mais tarde com outra modificação e então
% é agora inutil. Talvez seja demais adicionar à regra
%
Finally, these rules end up ensuring that a recursive ADT will deconstruct
itself once, and move the recursive argument from its constructor to the linear
context -- essentially forcing it to be used as a function argument, and most
usually that function will be the synthesis recursive call.
